Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-03-07 mpaper/paper.tex
Contour plots and other changes

\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\adj}{adj}
\DeclareMathOperator{\tr}{tr}

\DeclarePairedDelimiterX{\infdivx}[2]{(}{)}{%
  #1\;\delimsize\|\;#2%
}
\newcommand{\DKL}{D_{\mathrm{KL}}\infdivx}
\newcommand{\Eq}{\mathbb{E}_{(\mathbf{u}, \mathbf{r}) \sim \approximation}}
\newcommand{\V}{V_{\mathbf{r}}}
\newcommand{\dx}{\,d\mathbf{r}\,d\mathbf{u}}
\newcommand{\vbound}{\frac{\rinf + \log|\mathcal{A}|}{1 - \gamma}}
\newcommand{\rinf}{\lVert \mathbf{r} \rVert_\infty}

@@

\newcommand{\pfull}{p(\mathcal{D}, \mathbf{u}, \mathbf{r})}
\newcommand{\approximation}{q(\mathbf{u}, \mathbf{r})}

@@

\section{Introduction}
% problem description/statement

\section{Background}
% motivation, relevance (whatever that means)
% comprehensive literature survey

@@

\section{The WizWoz System}
% solution
% TODO: express all lambdas as exponentials (update the derivatives or sth)

@@

\[
  \begin{split}
    \left[ \frac{\partial \Kuu}{\partial \lambda_i} \right]_{j,k} =
    k_{\bm\lambda}(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
    \left( \vphantom{\frac{1}{2}} \right. &-\frac{1}{2}(x_{\mathbf{u},j,i} -
    x_{\mathbf{u},k,i})^2 \\
    &- \mathbbm{1}[j \ne k]\sigma^2 \left. \vphantom{\frac{1}{2}} \right)
  \end{split}
\]
otherwise.

@@

\begin{proof}
        \frac{\partial q(\mathbf{u})}{\partial m} &=
        q(\mathbf{u})\dm\left[-\frac{Q}{2}\right]
        \\
        &= -\frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu)\dm[\mathbf{u} -
        \bm\mu] \\
        &= \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu).
    \begin{align}
      \frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} &= \dB\left[\frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\exp \left( -\frac{Q}{2} \right)\right] \nonumber \\
                                                         &= \dB\left[\frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\right]\exp \left( -\frac{Q}{2} \right) \nonumber \\
                                                         &+ \frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\dB\left[\exp\left( -\frac{Q}{2} \right)\right] \nonumber \\
                                                         &= -\frac{1}{2} q(\mathbf{u}) \left( |\bm\Sigma|^{-3/2} \frac{\partial |\bm\Sigma|}{\partial \mathbf{B}} + \frac{\partial Q}{\partial \mathbf{B}} \right). \label{eq:2_before}
    \end{align}
    For derivatives with respect to $\bm\Sigma$, we can refer to Petersen and
    Pedersen \cite{petersen2008matrix}:
    \begin{equation} \label{eq:partial_derivatives1}
      \frac{\partial |\bm\Sigma|}{\partial \bm\Sigma} = |\bm\Sigma|\bm\Sigma^{-\intercal}, \quad \frac{\partial Q}{\partial \mathbf{B}} = -\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal},
    \end{equation}
    while we can use an online tool by Laue et
    al.\footnote{\url{http://www.matrixcalculus.org/}}
    \cite{DBLP:conf/nips/LaueMG18} for the remaining ones:
    \begin{equation} \label{eq:partial_derivatives2}
      \begin{gathered}
        \frac{\partial |\bm\Sigma|}{\partial \mathbf{B}} =
        (\adj(\mathbf{T}) + \adj(\bm\Sigma))\mathbf{B}, \\
        \dB[(\mathbf{u} - \bm\mu)^\intercal\bm\Sigma^{-1}(\mathbf{u} -
        \bm\mu)] = -(\mathbf{TUT} +
        \bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1})\mathbf{B}.
      \end{gathered}
    \end{equation}
    Substituting results from \eqref{eq:partial_derivatives1} and
    \eqref{eq:partial_derivatives2} back into \eqref{eq:2_before} gives:
    \begin{alignat*}{3}
      \frac{\partial q(\mathbf{u})}{\partial \bm\Sigma} &=
      \frac{1}{2}q(\mathbf{u})(&&\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal}
      - \bm\Sigma^{-\intercal}), \\
      \frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} &=
      \frac{1}{2}q(\mathbf{u}) \{&&|\bm\Sigma|^{-3/2}(\adj(\mathbf{T}) +
      \adj(\bm\Sigma)) \\
      & &&+ \mathbf{TUT} + \bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} \} \mathbf{B}.
    \end{alignat*}
  \item Using a result by Rasmussen and Williams % TODO: update the proof
                                % according to the new result
    \cite{DBLP:books/lib/RasmussenW06},
      \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} =
      q(\mathbf{r} \mid \mathbf{u}) \dl
      \left[-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} -
        \frac{1}{2}\log|\Kuu| \right] = \frac{1}{2}q(\mathbf{r} \mid \mathbf{u})\tr
      \left((\Kuu^{-1}\mathbf{u}\mathbf{u}^\intercal\Kuu^{-\intercal} -
        \Kuu^{-1}) \frac{\partial \Kuu}{\partial \lambda_i} \right).
    The remaining derivative is
      \frac{\partial \Kuu}{\partial \lambda_i} =
      \begin{cases}
        \frac{1}{\lambda_i}\Kuu & \text{if } i = 0, \\
        \Luu & \text{otherwise,}
      \end{cases}
    where
        [\Luu]_{j,k} &= \dl k_{\bm\lambda}(\mathbf{x}_{\mathbf{u},j},
        \mathbf{x}_{\mathbf{u},k}) \\
        &= k_{\bm\lambda}(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
        \dl \left[-\frac{1}{2}(\mathbf{x}_{\mathbf{u},j} -
          \mathbf{x}_{\mathbf{u},k})^\intercal \bm\Lambda
          (\mathbf{x}_{\mathbf{u},j} - \mathbf{x}_{\mathbf{u},k}) -
          \mathbbm{1}[j \ne k]\sigma^2\tr(\bm\Lambda) \right] \\
        &= k_{\bm\lambda}(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
        \dl \left[-\frac{1}{2}\sum_{l=1}^d \lambda_l
          (x_{\mathbf{u},j,l} - x_{\mathbf{u},k,l})^2 -
          \mathbbm{1}[j \ne k]\sigma^2\sum_{l=1}^d \lambda_l \right] \\
        &= k_{\bm\lambda}(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
        \left( -\frac{1}{2}(x_{\mathbf{u},j,i} -
        x_{\mathbf{u},k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2 \right).
\end{proof}

@@

\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \Eq \left[ \log \frac{\pfull}{\approximation}
    \right] \\
    &= \iint \approximation \log
    \frac{\pfull}{\approximation}\dx.
  \end{split}
\end{equation}

@@

\begin{proposition} \label{thm:measurability}
  MDP value functions $V(s) : \mathbb{R}^{|\mathcal{S}|} \to \mathbb{R}$ (for $s
  \in \mathcal{S}$) are Lebesgue measurable.
\end{proposition}
\begin{proof}
  For any reward vector $\mathbf{r} \in \mathbb{R}^{|\mathcal{S}|}$, the
  collection of converged value functions $\{ \V(s) \mid s \in
  \mathcal{S} \}$ satisfy
  \begin{equation} \label{eq:linear_mdp}
    \V(s) = \log \sum_{a \in \mathcal{A}}
    \exp\left( r(s) + \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s, a,
      s')\V(s') \right)
  \end{equation}
  for all $s \in \mathcal{S}$. Let $s_0 \in \mathcal{S}$ be an arbitrary state.
  In order to prove that $V(s_0)$ is measurable, it is enough to show that for
  any $\alpha \in \mathbb{R}$, the set
  \[
    \left\{ \mathbf{r} \in \mathbb{R}^{|\mathcal{S}|} \; \middle|
    \begin{split}
      \;&\V(s_0) \in (-\infty, \alpha); \\
      \;&\V(s) \in \mathbb{R} \text{ for all } s \in \mathcal{S} \setminus \{ s_0 \}; \\
      \;&\text{ Equation \ref{eq:linear_mdp} is satisfied by all } s \in
      \mathcal{S}
    \end{split}
    \right\}
  \]
  is measurable. Since this set can be constructed in Zermelo-Fraenkel set
  theory \emph{without} the axiom of choice, it is measurable
  \cite{herrlich2006axiom}, which proves that $V(s)$ is a measurable function
  for any $s \in \mathcal{S}$.
\end{proof}

>>>>

\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\adj}{adj}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator*{\argmax}{arg\,max}

\DeclarePairedDelimiterX{\infdivx}[2]{(}{)}{%
  #1\;\delimsize\|\;#2%
}
\newcommand{\DKL}{D_{\mathrm{KL}}\infdivx}
\newcommand{\V}{V_{\mathbf{r}}}
\newcommand{\dx}{\,d\mathbf{r}\,d\mathbf{u}}
\newcommand{\vbound}{\frac{\rinf + \log|\mathcal{A}|}{1 - \gamma}}
\newcommand{\rinf}{\lVert \mathbf{r} \rVert_\infty}

@@

\newcommand{\pfull}{p(\mathcal{D}, \mathbf{u}, \mathbf{r})}
\newcommand{\approximation}{q(\mathbf{u}, \mathbf{r})}
\newcommand{\posterior}{p(\mathbf{u}, \mathbf{r} \mid \mathcal{D})}

@@

% TODO: add info from slides/handout
\section{Introduction}
% problem description/statement

% TODO: will need to rewrite this to avoid plagiarism
Inverse reinforcement learning (IRL)---a problem proposed by Russell in
1998 \cite{DBLP:conf/colt/Russell98}---asks us to find a reward function for a
Markov decision process (MDP) that best explains a set of given demonstrations.
IRL is important because reward functions can be hard to define manually
\cite{DBLP:conf/icml/PieterN04,DBLP:journals/corr/abs-1806-06877}, and rewards
are not entirely specific to a given environment, allowing one to reuse the same
reward structure in previously unseen environments
\cite{DBLP:journals/corr/abs-1806-06877,DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}.
Moreover, IRL has seen a wide array of applications in autonomous vehicle
control \cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16} and
learning to predict another agent's behaviour
\cite{DBLP:journals/ai/BogertD18,DBLP:conf/aaai/VogelRGR12,ziebart2008maximum,DBLP:conf/huc/ZiebartMDB08,DBLP:conf/iros/ZiebartRGMPBHDS09}.
Most approaches in the literature (see Section \ref{sec:background}) make a
convenient yet unjustified assumption that the reward function can be expressed
as a linear combination of features. One proven way to abandon this assumption
is by representing the reward function as a Gaussian process (GP)
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11,DBLP:journals/corr/abs-1208-2112}.
The original approach used maximum likelihood estimation
\cite{DBLP:conf/nips/LevinePK11}, while the goal of this project is to use
variational inference (VI) instead, which learns approximate posterior
probability distributions instead of point estimates. This approach can prove
useful in three major ways:
\begin{enumerate}
\item Modelling full posterior distributions for various parameters can result
  in more precise reward predictions, as the model simply holds more
  information.
\item Having variance estimates for rewards can direct our choice in what data
  should be collected next.
\item An approximate Bayesian treatment of many parameters in the model guards
  against overfitting \cite{DBLP:conf/uai/JinDAS17}.
\end{enumerate}

\subsection{Statement of the Problem}

\begin{definition}[MDP]
  A \emph{Markov decision process} is a set $\mathcal{M} = \{ \mathcal{S},
  \mathcal{A}, \mathcal{T}, \gamma, r \}$, where $\mathcal{S}$ and
  $\mathcal{A}$ are sets of states and actions, respectively; $\mathcal{T} :
  \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is a function
  defined so that $\mathcal{T}(s, a, s')$ is the probability of moving to state $s'$
  after taking action $a$ in state $s$; $\gamma \in [0, 1)$ is the discount
  factor (with higher $\gamma$ values, it makes little difference whether a
  reward is received now or later, while with lower $\gamma$ values the future
  becomes gradually less and less important); and $r : \mathcal{S} \to
  \mathbb{R}$ is the reward function.
\end{definition}

In \emph{inverse reinforcement learning}, one is presented with an MDP without a
reward function $\mathcal{M} \setminus \{ r \}$ and a set of expert
demonstrations $\mathcal{D} = \{ \zeta_i \}_{i=1}^N$, where each demonstration
$\zeta_i = \{ (s_{i,0}, a_{i,0}), \dots, (s_{i,T}, a_{i,T}) \}$ is a multiset of
state-action pairs representing the actions taken by the expert during a
particular recorded session. Each state is also characterised by a number of
features. The goal of IRL is then to find $r$ such that the optimal policy under
$r$
\[ \pi^* = \argmax_\pi \mathbb{E}\left[ \left. \sum_{t=0}^\infty \gamma^t r(s_t)
    \right| \pi \right] \]
matches the actions in $\mathcal{D}$.

Following previous work on GP IRL
\cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}, we use a maximum
entropy IRL model \cite{ziebart2008maximum}, under which we have that
\[ P(a | s) \propto \exp(Q_{\mathbf{r}}(s, a)), \]
where
\[
  Q_{\mathbf{r}}(s, a) = r(s) + \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s, a, s')\V(s'),
\]
and $\V : \mathcal{S} \to \mathbb{R}$ is an MDP \emph{value function}, which
can be obtained by repeatedly applying the 'soft' version of the Bellman backup
operator until convergence:
\cite{DBLP:conf/nips/LevinePK11,supplementary_material}
\begin{equation} \label{eq:update_rule}
  \V(s) \coloneqq \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \gamma\sum_{s'
      \in \mathcal{S}} \mathcal{T}(s, a, s')\V(s') \right).
\end{equation}
The likelihood of the data can then be expressed as
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}
\begin{equation} \label{pDr}
  \begin{split}
    p(\mathcal{D} | r) &= \prod_{i=1}^N \prod_{t=1}^T p(a_{i,t} | s_{i,t}) \\
    &= \exp\left( \sum_{i=1}^N \sum_{t=1}^T Q_{\mathbf{r}}(s_{i,t}, a_{i,t}) - \V(s_{i,t}) \right).
  \end{split}
\end{equation}
However, a reward function learned by maximising this likelihood is not
transferable to new situations
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}. One needs to model the
reward structure in a way that would allow reward predictions for previously
unseen states.

One way to model rewards without assumptions of linearity is with a
\emph{Gaussian process} (GP). A GP is a collection of random variables, any
finite combination of which has a joint Gaussian distribution
\cite{DBLP:books/lib/RasmussenW06}. We write $r \sim \mathcal{GP}(0,
k)$ to say that $r$ is a GP with mean $0$ and covariance function
$k$, which uses a vector of hyperparameters $\bm\lambda$.
Covariance functions take two state feature vectors as input and quantify how
similar the two states are, in a sense that we would expect them to have similar
rewards.

As training a GP with $n$ data points has a time complexity of
$\mathcal{O}(n^3)$ \cite{DBLP:books/lib/RasmussenW06}, numerous approximation
methods have been suggested, many of which select a subset of data called
\emph{inducing points} and focus most of the training effort on them
\cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be the matrix of
features at inducing states, $\mathbf{u}$ the rewards at those states, and
$\mathbf{r}$ a vector with $r(\mathcal{S})$ as elements. Then the full joint
probability distribution can be factorised as
\[
  \pfull = p(\mathbf{u}) \times p(\mathbf{r} | \mathbf{u}) \times p(\mathcal{D} | r),
\]
where
\begin{align*}
  p(\mathbf{u}) &= \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu) \\
                &= \frac{1}{(2\pi)^{m/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2} \mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} \right) \\
                &= \exp\left(-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} - \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right)
\end{align*}
is the GP prior \cite{DBLP:books/lib/RasmussenW06}, where $m \in \mathbb{N}$ is
the number of inducing points. The GP posterior is a
multivariate Gaussian \cite{DBLP:conf/nips/LevinePK11}
\begin{equation} \label{eq:r}
  p(\mathbf{r} | \bm\lambda, \mathbf{X_u}, \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal\Kuu^{-1}\Kru),
\end{equation}
and $p(\mathcal{D} | r)$ is as in Equation \ref{pDr}. The matrices such as
$\Kru$ are called \emph{covariance matrices} and are defined as
$[\Kru]_{i,j} = k(\mathbf{x}_{\mathbf{r},i},
\mathbf{x}_{\mathbf{u},j})$, where $\mathbf{x}_{\mathbf{r},i}$ and
$\mathbf{x}_{\mathbf{u},j}$ denote feature vectors for the $i$th state in
$\mathcal{S}$ and the $j$th state in $\mathbf{X_u}$, respectively
\cite{DBLP:conf/uai/JinDAS17}.

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\bm\lambda$, inducing rewards $\mathbf{u}$, and the rewards for all relevant
states $\mathbf{r}$. While the previous paper to consider this IRL model
computed maximum likelihood (ML) estimates for $\bm\lambda$ and $\mathbf{u}$,
and made an assumption that $\mathbf{r}$ in Equation \ref{eq:r} has zero variance
\cite{DBLP:conf/nips/LevinePK11}, we aim to avoid this assumption and use
VI to approximate the full posterior distribution $\posterior$.
\emph{Variational inference} is an approximation technique for probability
densities \cite{blei2017variational}. Let $\approximation$ be our approximating
family of probability distributions for $\posterior$ with its own hyperparameter
vector $\bm\nu$. Then the job of VI algorithms is to optimise $\bm\nu$ in order
to minimise the \emph{Kullback-Leibler} (KL) divergence between the original
probability distribution and our approximation.  KL divergence (asymmetrically)
measures how different the two distributions are, and in this case can be
defined as \cite{blei2017variational}
\begin{align*}
  \DKL{\approximation}{\posterior} &= \mathbb{E}[\log\approximation - \log\posterior ] \\
                                   &= \mathbb{E}[\log\approximation - \log\pfull] \\
                                   &+ \mathbb{E}[\log p(\mathcal{D}, \mathbf{X_u})].
\end{align*}
The last term is both hard to compute and constant w.r.t. $\approximation$
\cite{blei2017variational}, so we can remove it from our optimisation objective.
The negation of what remains is often called the \emph{evidence lower bound}
(ELBO) and is defined as\footnote{Throughout the proposal, all integrals should
  be interpreted as definite integrals over the entire sample space.}
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \mathbb{E}\left[ \log \frac{\pfull}{\approximation} \right] \\
    &= \iiint \log \frac{\pfull}{\approximation} \approximation\dx.
  \end{split}
\end{equation}

By considering full probability distributions instead of point estimates---as
long as the approximations are able to capture important features of the
posterior---our predictions are likely to be more accurate and rely on fewer
assumptions. Moreover, we hope to make use of various recent advancements in VI
for both time complexity and approximation distribution fit (see Section
\ref{sec:background}), making the resulting algorithm competitive both in terms of
running time and model fit.

The project is primarily concerned with investigating how a VI formulation of
the GP IRL model compares against the original ML approach. Most importantly, we
aim to compare how the two algorithms converge both over time and as the number
of demonstrations increases. It would also be interesting to see how close the
approximation of the posterior distribution is to the real thing. Finally, it is
reasonable to conjecture that VI can outperform point estimates when dealing
with more uncertainty, e.g., when experts make mistakes. We can easily
investigate this by adapting how the evaluation data is generated.

\section{Background} \label{sec:background}
% motivation, relevance (whatever that means)
% comprehensive literature survey

@@

% TODO: how do we name this?
\section{The WizWoz System}
% solution
% TODO: express all lambdas as exponentials (update the derivatives or sth)

@@

\[
  \begin{split}
    \left[ \frac{\partial \Kuu}{\partial \lambda_i} \right]_{j,k} =
    k(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
    \left( \vphantom{\frac{1}{2}} \right. &-\frac{1}{2}(x_{\mathbf{u},j,i} -
    x_{\mathbf{u},k,i})^2 \\
    &- \mathbbm{1}[j \ne k]\sigma^2 \left. \vphantom{\frac{1}{2}} \right)
  \end{split}
\]
otherwise.

@@

%\begin{proof}
%  \leavevmode
%  \begin{enumerate}
%  \item
%    \[
%      \begin{split}
%        \frac{\partial q(\mathbf{u})}{\partial m} &=
%        q(\mathbf{u})\dm\left[-\frac{Q}{2}\right]
%        \\
%        &= -\frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
%        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu)\dm[\mathbf{u} -
%        \bm\mu] \\
%        &= \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
%        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu).
%      \end{split}
%    \]
%  \item
%    \begin{align}
%      \frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} &= \dB\left[\frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\exp \left( -\frac{Q}{2} \right)\right] \nonumber \\
%                                                         &= \dB\left[\frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\right]\exp \left( -\frac{Q}{2} \right) \nonumber \\
%                                                         &+ \frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\dB\left[\exp\left( -\frac{Q}{2} \right)\right] \nonumber \\
%                                                         &= -\frac{1}{2} q(\mathbf{u}) \left( |\bm\Sigma|^{-3/2} \frac{\partial |\bm\Sigma|}{\partial \mathbf{B}} + \frac{\partial Q}{\partial \mathbf{B}} \right). \label{eq:2_before}
%    \end{align}
%    For derivatives with respect to $\bm\Sigma$, we can refer to Petersen and
%    Pedersen \cite{petersen2008matrix}:
%    \begin{equation} \label{eq:partial_derivatives1}
%      \frac{\partial |\bm\Sigma|}{\partial \bm\Sigma} = |\bm\Sigma|\bm\Sigma^{-\intercal}, \quad \frac{\partial Q}{\partial \mathbf{B}} = -\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal},
%    \end{equation}
%    while we can use an online tool by Laue et
%    al.\footnote{\url{http://www.matrixcalculus.org/}}
%    \cite{DBLP:conf/nips/LaueMG18} for the remaining ones:
%    \begin{equation} \label{eq:partial_derivatives2}
%      \begin{gathered}
%        \frac{\partial |\bm\Sigma|}{\partial \mathbf{B}} =
%        (\adj(\mathbf{T}) + \adj(\bm\Sigma))\mathbf{B}, \\
%        \dB[(\mathbf{u} - \bm\mu)^\intercal\bm\Sigma^{-1}(\mathbf{u} -
%        \bm\mu)] = -(\mathbf{TUT} +
%        \bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1})\mathbf{B}.
%      \end{gathered}
%    \end{equation}
%    Substituting results from Equations \ref{eq:partial_derivatives1} and
%    \ref{eq:partial_derivatives2} back into Equation \ref{eq:2_before} gives:
%    \begin{alignat*}{3}
%      \frac{\partial q(\mathbf{u})}{\partial \bm\Sigma} &=
%      \frac{1}{2}q(\mathbf{u})(&&\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal}
%      - \bm\Sigma^{-\intercal}), \\
%      \frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} &=
%      \frac{1}{2}q(\mathbf{u}) \{&&|\bm\Sigma|^{-3/2}(\adj(\mathbf{T}) +
%      \adj(\bm\Sigma)) \\
%      & &&+ \mathbf{TUT} + \bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} \} \mathbf{B}.
%    \end{alignat*}
%  \item Using a result by Rasmussen and Williams % TODO: update the proof
%                                % according to the new result
%    \cite{DBLP:books/lib/RasmussenW06},
%    \[
%      \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} =
%      q(\mathbf{r} \mid \mathbf{u}) \dl
%      \left[-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} -
%        \frac{1}{2}\log|\Kuu| \right] = \frac{1}{2}q(\mathbf{r} \mid \mathbf{u})\tr
%      \left((\Kuu^{-1}\mathbf{u}\mathbf{u}^\intercal\Kuu^{-\intercal} -
%        \Kuu^{-1}) \frac{\partial \Kuu}{\partial \lambda_i} \right).
%    \]
%    The remaining derivative is
%    \[
%      \frac{\partial \Kuu}{\partial \lambda_i} =
%      \begin{cases}
%        \frac{1}{\lambda_i}\Kuu & \text{if } i = 0, \\
%        \Luu & \text{otherwise,}
%      \end{cases}
%    \]
%    where
%    \[
%      \begin{split}
%        [\Luu]_{j,k} &= \dl k(\mathbf{x}_{\mathbf{u},j},
%        \mathbf{x}_{\mathbf{u},k}) \\
%        &= k(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
%        \dl \left[-\frac{1}{2}(\mathbf{x}_{\mathbf{u},j} -
%          \mathbf{x}_{\mathbf{u},k})^\intercal \bm\Lambda
%          (\mathbf{x}_{\mathbf{u},j} - \mathbf{x}_{\mathbf{u},k}) -
%          \mathbbm{1}[j \ne k]\sigma^2\tr(\bm\Lambda) \right] \\
%        &= k(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
%        \dl \left[-\frac{1}{2}\sum_{l=1}^d \lambda_l
%          (x_{\mathbf{u},j,l} - x_{\mathbf{u},k,l})^2 -
%          \mathbbm{1}[j \ne k]\sigma^2\sum_{l=1}^d \lambda_l \right] \\
%        &= k(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
%        \left( -\frac{1}{2}(x_{\mathbf{u},j,i} -
%        x_{\mathbf{u},k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2 \right).
%      \end{split}
%    \]
%  \end{enumerate}
%\end{proof}

@@

@@

\begin{proposition} \label{thm:measurability}
  MDP value functions $V(s) : \mathbb{R}^{|\mathcal{S}|} \to \mathbb{R}$ (for $s
  \in \mathcal{S}$) are Lebesgue measurable.
\end{proposition}
\begin{proof}
  For any reward vector $\mathbf{r} \in \mathbb{R}^{|\mathcal{S}|}$, the
  collection of converged value functions $\{ \V(s) \mid s \in
  \mathcal{S} \}$ satisfy
  \begin{equation} \label{eq:linear_mdp}
    \V(s) = \log \sum_{a \in \mathcal{A}}
    \exp\left( r(s) + \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s, a,
      s')\V(s') \right)
  \end{equation}
  for all $s \in \mathcal{S}$. Let $s_0 \in \mathcal{S}$ be an arbitrary state.
  In order to prove that $V(s_0)$ is measurable, it is enough to show that for
  any $\alpha \in \mathbb{R}$, the set
  \[
    \left\{ \mathbf{r} \in \mathbb{R}^{|\mathcal{S}|} \; \middle|
    \begin{aligned}
      \;&\V(s_0) \in (-\infty, \alpha); \\
      \;&\V(s) \in \mathbb{R} \text{ for all } s \in \mathcal{S} \setminus \{ s_0 \}; \\
      \;&\text{ Equation \ref{eq:linear_mdp} is satisfied by all } s \in
      \mathcal{S}
    \end{aligned}
    \right\}
  \]
  is measurable. Since this set can be constructed in Zermelo-Fraenkel set
  theory \emph{without} the axiom of choice, it is measurable
  \cite{herrlich2006axiom}, which proves that $V(s)$ is a measurable function
  for any $s \in \mathcal{S}$.
\end{proof}

\begin{proposition} \label{thm:bound}
  If the initial values of the MDP value function satisfy the following
  bound, then the bound remains satisfied throughout value iteration:
  \begin{equation} \label{eq:bound}
    |\V(s)| \le \vbound.
  \end{equation}
\end{proposition}
\begin{proof}
  We begin by considering Equation \ref{eq:bound} without taking the absolute value of
  $\V(s)$, i.e.,
  \begin{equation} \label{eq:positive_bound}
    \V(s) \le \vbound,
  \end{equation}
  and assuming that the initial values of $\{ \V(s) \mid s \in
  \mathcal{S} \}$ already satisfy Equation \ref{eq:positive_bound}. Recall that for
  each $s \in \mathcal{S}$, the value of $\V(s)$ is updated by applying Equation
  \ref{eq:update_rule}. Note that both $\log$ and $\exp$ are increasing
  functions, $\gamma > 0$, and the $\mathcal{T}$ function gives a probability (a
  non-negative number). Thus
  \begin{align*}
    \V(s) &\le \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s, a, s')\frac{\rinf + \log|\mathcal{A}|}{1 - \gamma} \right) \\
          &= \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \frac{\gamma (\rinf + \log|\mathcal{A}|)}{1 - \gamma}\sum_{s' \in \mathcal{S}} \mathcal{T}(s, a, s') \right) \\
          &= \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \frac{\gamma (\rinf + \log|\mathcal{A}|)}{1 - \gamma} \right)
  \end{align*}
  by the definition of $\mathcal{T}$. Then
  \begin{align*}
    \V(s) &\le \log \left( |\mathcal{A}| \exp\left( r(s) + \frac{\gamma (\rinf + \log|\mathcal{A}|)}{1 - \gamma} \right) \right) \\
          &= \log \left( \exp\left( \log|\mathcal{A}| + r(s) + \frac{\gamma (\rinf + \log|\mathcal{A}|)}{1 - \gamma} \right) \right) \\
          &= \log|\mathcal{A}| + r(s) + \frac{\gamma (\rinf + \log|\mathcal{A}|)}{1 - \gamma} \\
          &= \frac{\gamma (\rinf + \log|\mathcal{A}|) + (1 - \gamma)(\log|\mathcal{A}| + r(s))}{1 - \gamma} \\
          &\le \frac{\gamma (\rinf + \log|\mathcal{A}|) + (1 - \gamma)(\log|\mathcal{A}| + \rinf)}{1 - \gamma} \\
          &= \vbound
  \end{align*}
  by the definition of $\rinf$.

  The proof for
  \begin{equation} \label{eq:negative_bound}
    \V(s) \ge \frac{\rinf + \log|\mathcal{A}|}{\gamma - 1}
  \end{equation}
  follows the same argument until we get to
  \begin{align*}
    \V(s) &\ge \frac{\gamma(\rinf + \log|\mathcal{A}|) + (\gamma - 1)(\log|\mathcal{A}| + r(s))}{\gamma - 1} \\
          &\ge \frac{\gamma(\rinf + \log|\mathcal{A}|) + (\gamma - 1)(-\log|\mathcal{A}| -\rinf)}{\gamma - 1} \\
          &= \frac{\rinf + \log|\mathcal{A}|}{\gamma - 1},
  \end{align*}
  where we use the fact that $r(s) \ge -\rinf - 2\log|\mathcal{A}|$. Combining
  Equations \ref{eq:positive_bound} and \ref{eq:negative_bound} gives
  Equation \ref{eq:bound}.
\end{proof}
